The Design, Semantics, and Implementation of the Synchronous Reactive Language \CEU

### 2015-10-10 paper.tex
Novo titulo Nova intro   - complexidades de Esterel Revisoes Roberto

% Title portion
\title{The Design, Semantics, and Implementation of \CEU: a Synchronous Reactive Language}

@@

An established alternative to \C in the field of embedded systems is the family 
of reactive synchronous languages \cite{rp.twelve}.
%
Two major styles of synchronous languages have evolved:
in the \emph{control}--\emph{imperative} style, programs are structured with 
control flow primitives, such as parallelism, repetition, and preemption;
in the \emph{dataflow}--\emph{declarative} style, programs can be seen as 
graphs of values, in which a change to a value is propagated through its 
dependencies without explicit programming.
%
Considering the control-based languages, Esterel~\cite{esterel.ieee91} was the 
first to appear and influenced a number of embedded languages, such as 
\emph{Reactive-C}~\cite{rp.rc}, \emph{OSM}~\cite{wsn.osm}, 
\emph{Sync-C}~\cite{rp.synchc}, and \emph{PRET-C}~\cite{rp.pretc}.

\CEU is another Esterel-based language targeting embedded and (soft) real-time 
systems with novel functionalities:
%
\begin{itemize}
%
\item Stack-based execution for internal events, which provide a limited form 
of coroutines.
%
\item A static temporal analysis and deterministic execution semantics that 
allows programs to safely share memory.
%
\item A finalization mechanism for safe abortion of lines of execution holding 
external resources.
%
\item First-class synchronized timers.
\end{itemize}

In this work, we discuss the design of \CEU and propose a formal semantics for 
a small synchronous kernel that represents a subset of the language covering 
these new functionalities.
%
We also present an implementation of \CEU with two back ends:
one aiming for resource efficiency and interoperability with \C,
and another based on a virtual machine that allows remote reprogramming.
%
Our implementations target resource-constrained devices, such as \emph{Arduino} 
and \emph{MICAz} sensor nodes based on 8-bit microcontrollers.%
\footnote{
Arduino: \url{https://www.arduino.cc/en/Main/arduinoBoardUno}
\\
MICAz: \url{http://www.memsic.com/userfiles/files/Datasheets/WSN/micaz_datasheet-t.pdf}
\\
(Both use the \emph{ATmega328} microcontroller with 32 Kbytes of FLASH and 2 
Kbytes of SRAM.)}
%:\url{http://www.atmel.com/devices/atmega328.aspx}}

In previous work~\cite{ceu.sensys13,ceu.terra}, we employed \CEU in the context 
of wireless sensor networks, developing a number of applications, protocols, 
and drivers.
%
We evaluated the expressiveness of \CEU in comparison to event-driven code in 
\C and attested a reduction in source code size (around 25\%) with a small 
increase in memory usage (around 5--10\% for \emph{text} and 
\emph{data})~\cite{ceu.sensys13}.
%
For the \VM back end, a simple application that blinks three LEDs periodically 
occupies less than 100 bytes and can be completely transmitted in 4 radio 
messages.
In contrast, the full image of a \C-compiled version occupies more than 2 
Kbytes~\cite{ceu.terra}.%
\footnote{
The \VM is preloaded in the devices with some \emph{I/O} functionality, whereas 
\C-based binaries are full images.
}

@@

\CEU is a synchronous reactive language inspired by Esterel with support for 
multiple concurrent lines of execution known as \emph{trails}.
By reactive, we mean that programs are stimulated by the environment through 
input events that are broadcast to all awaiting trails.
By synchronous, we mean that all trails at any given moment are either reacting 
to the current event or are awaiting another event;
in other words, trails are never reacting to different events.

@@

Regarding the similarities, Figure~\ref{lst.abro} shows side-by-side the 
implementations in Esterel and \CEU for the following control 
specification~\cite{esterel.primer}:
%
\emph{``Emit an output O as soon as two inputs A and B have occurred.
Reset this behavior each time the input R occurs''.}
%
The first phrase of the specification, awaiting and emitting the events, is 
translated almost identically in the two languages (ln. 4--9, in both 
implementations), given that Esterel's `$\|$' and \CEU's \code{par/and} 
constructs are equivalent.
%
For the second phrase, the reset behavior, the Esterel version uses a 
\code{abort-when} (ln. 3--10), which serves the same purpose of \CEU's 
\code{par/or} (ln. 3--12):
the occurrence of event \code{R} aborts the awaiting statements in parallel and 
restarts the enclosing \code{loop}.

\begin{figure}[b]
\begin{minipage}[t]{0.48\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3.5em,mathescape=true]
// ESTEREL
loop
   abort
      [
         await A
      $\|$
         await B
      ];
      emit O
   when R
end
\end{lstlisting}
\end{minipage}
%
\begin{minipage}[t]{0.48\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3.5em]
// CEU
loop do
   par/or do
      par/and do
         await A;
      with
         await B;
      end
      emit O;
   with
      await R;
   end
end
\end{lstlisting}
\end{minipage}
%\rule{8.4cm}{0.37pt}
\caption{ A control specification implemented in Esterel and \CEU:
\emph{``Emit \code{O} after \code{A} and \code{B}, resetting each \code{R}.''}
\label{lst.abro}
}
\end{figure}

\CEU has a strong imperative flavor, with explicit control flow through 
sequences, loops, parallels, and also assignments.
Being designed for control-intensive applications, it provides support for 
concurrent lines of execution and broadcast communication through events.
%
In addition, \CEU also employs the synchronous model, in which programs advance 
in a sequence of discrete reactions to external events.
Internal computations within a reaction (e.g. expressions, assignments, and 
system calls) are considered to take no time in accordance with the synchronous 
hypothesis~\cite{rp.hypothesis}.
The \code{await} statements are the only ones that halt a running reaction and 
allow a program to advance in this notion of time.
%
To ensure that reactions run in bounded time and programs always progress, 
loops are statically required to contain at least one \code{await} statement in 
all possible paths~\cite{ceu.sensys13,esterel.primer}.
%
\CEU shares the same limitations with Esterel and synchronous languages in 
general:
computations that run in unbounded time (e.g., cryptography, image processing) 
do not fit the zero-delay hypothesis, and cannot be elegantly implemented.

@@

Esterel makes no semantic distinctions between internal and external signals, 
both having only the notion of either presence or absence during an entire 
reaction~\cite{esterel.preemption}.
%
In \CEU, external input events are unique within reactions and programs cannot 
emit them, resulting in an intrinsic queue-based handling.
%
In contrast, programs can emit internal events but these follow a stack-based 
execution policy, similar to subroutine calls in typical programming languages.
%
Figure~\ref{lst.prints} illustrates the use of internal signals (events) in 
Esterel and \CEU.
%
In the version in Esterel, when \code{A} occurs, \code{B} is emitted (ln.  
5--6) and both events become active, resulting in the invocation of \code{f()} 
and \code{g()} in no particular order.
%
In the version in \CEU, the occurrence of \code{A} makes the program behave as 
follows (with the stack contents in italics):
%
{\small
\begin{enumerate}
\setlength{\itemsep}{0pt}
\item 1st trail awakes (ln. 5), emits \code{b}, and pauses.\\
    \emph{stack: [1st-trail]}
\item 2nd trail awakes (ln. 9), calls \code{\_g()}, and terminates.\\
    \emph{stack: [1st-trail]}
\item 1st trail (on top of the stack) resumes, calls \code{\_f()}, and 
    terminates.\\
    \emph{stack: []}
\item Both trails have terminated, so the \code{par/and} rejoins, and the 
program also terminates;
\end{enumerate}
}

@@

Another distinction regarding event handling in comparison to \CEU is that 
Esterel supports same-cycle bi-directional 
communication~\cite{esterel.compiling}, i.e., two threads can react to one 
another during the same cycle due to mutual signal dependency.
%
\CEU imposes a restriction that an \code{await} is only valid for the next 
reaction, i.e., if an \code{await} and \code{emit} occur simultaneously in 
parallel trails, the \code{await} does not awake.
%
These \emph{delayed awaits} avoid corner cases of instantaneous termination and 
re-execution of statements in the same reaction (known as \emph{schizophrenic 
statements}~\cite{esterel.schizo,esterel.schizo2}).
%
The example that follows relies on this restriction to avoid infinite 
execution:

\begin{lstlisting}[numbers=left,xleftmargin=2em]
event void e,f;
loop do
    par/or do
        await e;
    with
        emit e;     // w/o the restriction, the emit awakes 1st trail
        await f;    // and restarts the loop instantaneously
    end
end
\end{lstlisting}

Both sides of the \code{par/or} have an \code{await} statement to avoid 
instantaneous termination (ln. 4,7).
However, if the \code{emit} (ln. 6) could awake the \code{await} (ln. 4) in the 
same reaction that reaches them, the \code{par/or} would terminate and restart 
the \code{loop} instantaneously, resulting in infinite execution.

In atypical scenarios requiring immediate awake, delayed awaits can be 
circumvented by manually copying or transforming the code to execute on awake.
%
From our experience, in some cases we need to execute a block of code 
periodically from internal event requests, \emph{including the current 
reaction}, as illustrated in the left of Figure~\ref{lst.delay}.
In this case, moving the \code{await} to the end of the loop (ln. 10) makes the 
periodic code to also execute immediately (ln. 9), and then in reactions to 
each \code{emit} request (ln. 5).
If the periodic \code{emit} depends on a condition, as illustrated in the right 
of Figure~\ref{lst.delay}, the code becomes more intricate because we need a 
state variable (ln. 2) and also to copy the condition test to the periodic code 
(ln. 13).
%
On the one hand, we transfer the burden of dealing with these corner cases to 
the programmer.
On the other hand, we simplify the semantics of the language and eliminate the 
need for analysis to deal with schizophrenic statements.

\begin{figure}
\begin{minipage}[t]{0.50\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3em]
event void e;
par do
    loop do
        <...>     // code that awaits
        emit e;   // periodic request
    end
with
    loop do
        <...>     // code to execute
        await e;  // await after
    end
end
\end{lstlisting}
\end{minipage}
%
\begin{minipage}[t]{0.50\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3em]
event void e;
var bool should_execute = false;
par do
    loop do
        <...>         // code that awaits
        if <...> then // some conditon
            should_execute = true;
            emit e;
        end
    end
with
    loop do
        if should_execute then
            <...>     // code to execute
        end
        await e;
    end
end
\end{lstlisting}
\end{minipage}
%
\caption{ Examples that circumvent the \emph{delayed await} restriction by 
post-fixing the \code{await} inside the \code{loop} (in the left), and by 
copying the condition test (in the right).
\label{lst.delay}
}
\end{figure}

@@

Embedded applications make extensive use of global memory, such as for 
accessing resources through memory-mapped registers.
Hence, an important goal of \CEU is to ensure a reliable behavior for programs 
with concurrent lines of execution sharing memory.
%
Esterel is only deterministic with respect to reactive control: ``the same 
sequence of inputs always produces the same sequence of 
outputs''~\cite{esterel.primer}.
%
However, the execution order for operations with side-effects within a reaction 
is non-deterministic: ``if there is no control dependency, as in \code{<<call 
f1() || call f2()>>}, the order is unspecified and it would be an error to rely 
on it''~\cite{esterel.primer}.
%
Therefore, Esterel forbids sharing memory between lines of execution: 
\emph{``if a variable is written by some thread, then it can neither be read 
nor be written by concurrent threads''}~\cite{esterel.primer}.

Concurrency in \CEU is characterized when two or more trail segments in 
parallel execute during the same reaction.
A trail segment is a sequence of statements followed by an \code{await} (or 
termination).
%
In the program in the left of Figure~\ref{lst.shared}, the two assignments to 
\code{x} (ln. 5,8) can never run concurrently, because each trail segment 
reacts to a different input event (ln. 4,7) and, according to the semantics of 
\CEU, cannot occur simultaneously.
%
However, the program in the right is non-deterministic, because the two 
assignments to \code{y} (ln. 5,8) occur in the same reaction to input \code{A} 
(ln. 4,7).

\begin{figure}[b]
\begin{minipage}[t]{0.50\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3em]
input void A, B;
var int x = 1;
par/and do
    await A;
    x = x + 1;
with
    await B;
    x = x * 2;
end
\end{lstlisting}
\end{minipage}
%
\begin{minipage}[t]{0.50\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3em]
input void A;
var int y = 1;
par/and do
    await A;
    y = y + 1;
with
    await A;
    y = y * 2;
end
\end{lstlisting}
\end{minipage}
%\rule{8.4cm}{0.37pt}
\caption{ Shared-memory concurrency in \CEU:
the code in the left is safe because the trails access \code{x} atomically in 
different reactions;
the code in the right is unsafe because both trails access \code{y} in the same 
reaction.
\label{lst.shared}
}
\end{figure}

\CEU performs a temporal analysis at compile time and detects concurrent 
accesses to shared variables, as follows:
\emph{if a variable is written in a trail segment, then a concurrent trail 
segment cannot read or write to that variable, nor dereference a pointer of 
that variable type.}
An analogous policy is applied for pointers \emph{vs} variables and pointers 
\emph{vs} pointers, as well as for system calls with side effects (e.g., 
\code{printf}).
%The algorithm for the analysis holds the set of all events in preceding 
%\code{await} statements for each variable access.
%Then, the sets for all accesses in parallel trails are compared to assert that 
%no events are shared among them.
%Otherwise the compiler warns about the suspicious accesses.

@@

\begin{figure}
\begin{minipage}[t]{0.48\linewidth}
\begin{lstlisting}[mathescape=true]
// ESTEREL
input A;
[
    await A;
    call f1();
$\|$
    await A;
    call f2();
];
\end{lstlisting}
\end{minipage}
%
\begin{minipage}[t]{0.48\linewidth}
\begin{lstlisting}
// CEU
input void A;
par/and do
    await A;
    _f1();
with
    await A;
    _f2();
end
\end{lstlisting}
\end{minipage}
%\rule{8.4cm}{0.37pt}
\caption{ In Esterel, the execution order between \code{f1} and \code{f2} is 
unspecified, whereas in \CEU, \code{\_f1} executes before \code{\_f2} due to 
deterministic scheduling based on lexical order.
\label{lst.det}
}
\end{figure}

@@

To illustrate threats related to abortion, consider the unsafe example in the 
left of Figure~\ref{lst.fin}, which does not compile in \CEU.
It describes the state machine of a data collection protocol for sensor 
networks~\cite{wsn.ctp,ceu.sensys13}.
%
The input and output events represent the external interface of the protocol 
(ln. 1--2).
%
The protocol has to transmit a packet every minute (with \code{SEND\_ENQUEUE}), 
unless it receives a \code{RETRANSMIT} request to immediately re-transmit it, 
or a \code{STOP} request to terminate.
%
The protocol is composed of two main trails:
one simply monitors the stopping event (ln. 4);
the other periodically transmits the packet (ln. 6--19).
%
The periodic transmission is a loop that starts two other trails (ln. 7--18):
one handles the immediate retransmission request (ln. 8);
the other transmits the packet and waits for a confirmation (ln. 10--17).
%
%
The actual transmission (ln. 13--16) is enclosed with a \code{par/and} that 
takes at least one minute before looping (ln. 11), in accordance with the 
specification.
Note that the \code{emit SEND\_ENQUEUE} (ln. 15) is \emph{asynchronous}, 
handing to the radio driver a pointer to the lexically-scoped packet (ln. 13).
The driver makes the transmission in the background, holding the packet until 
it signals the application with the \code{SEND\_ACK} to acknowledge completion 
(ln. 16).
%
At any time, the client may request a retransmission (ln. 8), which terminates 
the \code{par/or} (ln. 7), aborts the ongoing transmission (ln. 15, if not 
idle), and restarts the loop (ln. 6).
%
The client may also request to stop the whole protocol at any time (ln. 4).
%
Therefore, if the sending trail is aborted by the \code{STOP} or 
\code{RETRANSMIT} requests, the packet buffer goes out of scope (ln. 13), 
leaving behind a \emph{dangling pointer} in the radio driver, which will 
possibly transmit corrupted data.
%
%\subsubsection{Finalization}
%\label{sec.ceu.fin}

The unsafe example in the left of Figure~\ref{lst.fin} does not compile because 
\CEU tracks the interaction of \code{par/or} compositions with local variables 
and stateful output events calls in order to preserve safe abortion of trails.
%
%Finalization clauses are fundamental to preserve the orthogonality of 
%\code{par/or} compositions in SSRP.
%
In fact, \CEU enforces the programmer to write a \emph{finalization} clause to 
accompany the output request.
The code in the right of Figure~\ref{lst.fin} properly cancels the packet 
transmission when the block of \code{buffer} goes out of scope, i.e., the 
finalization clause (after the \code{with}, ln. 18) executes automatically on 
external abortion.%
\footnote{
Note that the compiler only enforces the programmer to write the finalization 
clause, but cannot check if it actually handles the resource properly.
}
%
%Note also that \CEU environments rely on \C libraries that only provide 
%asynchronous I/O and non-blocking functions~\cite{ceu.sensys13}.

@@

\begin{itemize}
\item The program is awaiting in all trails, i.e., function $pop$ returns 
$(0,\{\})$.
\item The program terminates, i.e., the small-step rules transform the whole 
program into a $mem(id)$.
\end{itemize}
%
In Section~\ref{XXX} we show that by imposing syntactic restrictions to 
programs, reaction chains always reach one of these conditions in a finite 
number of steps, meaning that reactions to the environment always execute in 
bounded time.

@@

\subsection{The External \C API}

>>>>

% Title portion
\title{The Design, Semantics, and Implementation of the Synchronous Reactive 
Language \CEU}

@@

An established alternative to \C in the field of embedded systems is the family 
of reactive synchronous languages \cite{rp.twelve}.
%
Two major styles of synchronous languages have evolved:
in the \emph{control}--\emph{imperative} style, programs are structured with 
control flow primitives, such as parallelism, repetition, and preemption;
in the \emph{dataflow}--\emph{declarative} style, programs can be seen as 
graphs of values, in which a change to a value is propagated through its 
dependencies without explicit programming.
%
Considering the control-based languages, Esterel~\cite{esterel.ieee91} was the 
first to appear and succeed, influencing a number of embedded languages, such 
as \emph{Reactive-C}~\cite{rp.rc}, \emph{OSM}~\cite{wsn.osm}, 
\emph{Sync-C}~\cite{rp.synchc}, and \emph{PRET-C}~\cite{rp.pretc}.

Despite its success and influence, Esterel has a overly complex semantics that 
requires ingenious static analysis to detect and refuse programs with 
\emph{causality} and \emph{schizophrenia} 
problems~\cite{esterel.constructive,esterel.d1,esterel.d2,esterel.d3,esterel.d4,esterel.d5,esterel.d6,esterel.d7,esterel.d8}.
%
The complex semantics not only challenges the analysis and compilation of 
programs, but also results in incompatible or non-compliant implementations.
%
More importantly, it also affects the programmer's understanding capacity, 
which, after all, have to solve the errors when facing corner cases in 
programs.
%
Another drawback of the Esterel semantics consists of non-deterministic 
execution for intra-reaction statements, which prevents threads to share memory 
and interact with side-effect and stateful system calls.

In this work, we present \CEU, a new programming language that inherits the 
synchronous and imperative mindset of Esterel but diverges in fundamental 
semantic aspects.
%
Overall, \CEU has a simple semantics with fine-grained execution control, and a 
straightforward implementation targeting resource-constrained systems.
%
The list that follows summarizes the semantic peculiarities of \CEU:
%
\begin{itemize}
%
\item Stack-based execution for internal events, which provide a limited form 
of coroutines.
%
\item A static temporal analysis and deterministic execution semantics that 
allows programs to safely share memory.
%
\item A finalization mechanism for safe abortion of lines of execution holding 
external resources.
%
\item First-class synchronized timers.
\end{itemize}
%
We discuss the design of \CEU and present a formal semantics for a small 
synchronous kernel that represents a subset of the language covering these new 
functionalities.
%
We also present a lightweight implementation of \CEU with two back ends:
one aiming for resource efficiency and interoperability with \C,
and another based on a virtual machine that allows remote reprogramming.
%
Our implementations target resource-constrained devices, such as \emph{Arduino} 
and \emph{MICAz} sensor nodes based on 8-bit microcontrollers, showing that the 
peculiarities in the semantics of \CEU does not pose practical obstacles.%
\footnote{
Arduino: \url{https://www.arduino.cc/en/Main/arduinoBoardUno}
\\
MICAz: \url{http://www.memsic.com/userfiles/files/Datasheets/WSN/micaz_datasheet-t.pdf}
\\
(Both use the \emph{ATmega328} microcontroller with 32 Kbytes of FLASH and 2 
Kbytes of SRAM.)}
%:\url{http://www.atmel.com/devices/atmega328.aspx}}

In previous work~\cite{ceu.sensys13,ceu.terra}, we employed \CEU in the context 
of wireless sensor networks, developing a number of applications, protocols, 
and drivers.
%
We evaluated the expressiveness of \CEU in comparison to event-driven code in 
\C and attested a reduction in source code size (around 25\%) with a small 
increase in memory usage (around 5--10\% for \emph{text} and 
\emph{data})~\cite{ceu.sensys13}.
%
For the \VM back end, a simple application that blinks three LEDs periodically 
occupies less than 100 bytes and can be completely transmitted in 4 radio 
messages.

@@

\CEU is a synchronous reactive language inspired by Esterel with support for 
multiple concurrent lines of execution known as \emph{trails}.
By reactive, we mean that programs are stimulated by the environment through 
input events, which are broadcast to all awaiting trails.
By synchronous, we mean that trails at any given moment are either reacting to 
the current event or are awaiting another event;
in other words, trails never react to different events simultaneously.

@@

Regarding the similarities, Figure~\ref{lst.abro} shows side-by-side the 
implementations in Esterel and \CEU for the following control 
specification~\cite{esterel.primer}:
%
\emph{``Emit an output O as soon as two inputs A and B have occurred.
Reset this behavior each time the input R occurs''.}
%
The first phrase of the specification, awaiting and emitting the events, is 
translated almost identically in the two languages (ln. 4--9, in both 
implementations), given that Esterel's `$\|$' and \CEU's \code{par/and} 
constructs are equivalent.
%
For the second phrase, the reset behavior, the Esterel version uses a 
\code{abort-when} (ln. 3--10), which, in this case, serves the same purpose of 
\CEU's \code{par/or} (ln. 3--12):
the occurrence of event \code{R} aborts the awaiting statements in parallel and 
restarts the enclosing \code{loop}.

\begin{figure}
\begin{minipage}[t]{0.48\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3.5em,mathescape=true]
// ESTEREL
loop
   abort
      [
         await A
      $\|$
         await B
      ];
      emit O
   when R
end
\end{lstlisting}
\end{minipage}
%
\begin{minipage}[t]{0.48\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3.5em]
// CEU
loop do
   par/or do
      par/and do
         await A;
      with
         await B;
      end
      emit O;
   with
      await R;
   end
end
\end{lstlisting}
\end{minipage}
%\rule{8.4cm}{0.37pt}
\caption{ A control specification implemented in Esterel and \CEU:
\emph{``Emit \code{O} after \code{A} and \code{B}, resetting each \code{R}.''}
\label{lst.abro}
}
\end{figure}

\CEU employs the synchronous model, in which programs advance in a sequence of 
discrete reactions to external events.
%
It also has a strong imperative flavor, with explicit control flow through 
sequences, loops, and parallels, and also assignments.
%
Being designed for control-intensive applications, \CEU provides support for 
concurrent lines of execution and broadcast communication through events.
%
Internal computations within a reaction (e.g. expressions, assignments, and 
system calls) are considered to take no time in accordance with the synchronous 
hypothesis~\cite{rp.hypothesis}.
An \code{await} is the only statement that halts a running reaction and allows 
a program to advance in this notion of time.
%
To ensure that reactions run in bounded time and programs always progress, 
loops are statically required to contain at least one \code{await} statement in 
all possible paths~\cite{ceu.sensys13,esterel.primer}.
%
\CEU shares the same limitations with Esterel and synchronous languages in 
general:
computations that run in unbounded time (e.g., cryptography, image processing) 
do not fit the zero-delay hypothesis, and cannot be directly implemented.

@@

Esterel makes no semantic distinctions between internal and external signals.
In particular, programs can emit multiple external input signals, which may 
coexist during a reaction.
%
In \CEU, a reaction starts from the occurrence of an external input event and 
programs cannot emit inputs.
Therefore, the occurring input event is unique during the entire reaction, 
resulting in intrinsic queue-based handling.
%
In contrast, programs can emit internal events but these follow a stack-based 
execution policy, similar to subroutine calls in typical programming languages.
%
Figure~\ref{lst.prints} illustrates the use of internal signals (events) in 
Esterel and \CEU.
%
In Esterel, when \code{A} occurs, \code{B} is emitted (ln.  5--6) and both 
events become active, resulting in the invocation of \code{f()} and \code{g()} 
in no particular order.
%
In \CEU, the occurrence of \code{A} makes the program behave as follows:
%
{\small
\begin{enumerate}
\setlength{\itemsep}{0pt}
\item 1st trail awakes (ln. 5), emits \code{b}, and pauses.
\item 2nd trail awakes (ln. 9), calls \code{\_g()}, and terminates.
\item 1st trail (on top of the stack) resumes, calls \code{\_f()}, and 
    terminates.
\item Both trails have terminated, so the \code{par/and} rejoins, and the 
program also terminates.
\end{enumerate}
}

@@

Another distinction regarding event handling in comparison to \CEU is that 
Esterel supports same-cycle bi-directional 
communication~\cite{esterel.compiling}, i.e., two threads can react to one 
another during the same cycle due to mutual signal dependency.
%
\CEU has a different take, posing a tradeoff that an \code{await} is only valid 
for the next reaction, i.e., if an \code{await} and \code{emit} occur 
simultaneously in parallel trails, the \code{await} does not awake.
%
These \emph{delayed awaits} avoid corner cases of instantaneous termination and 
re-execution of statements in the same reaction (known as \emph{schizophrenic 
statements}~\cite{esterel.schizo,esterel.schizo2}).
%
The example in Figure~\ref{lst.await} illustrates this distinction, which 
prevents infinite execution by design.
Both sides of the \code{par/or} have an \code{await} statement to avoid 
instantaneous termination (ln. 4,7).
However, if the \code{emit} (ln. 6) could awake the \code{await} (ln. 4) in the 
same reaction that reaches them, the \code{par/or} would terminate and restart 
the \code{loop} instantaneously, resulting in infinite execution.

\begin{figure}
\begin{lstlisting}[numbers=left,xleftmargin=2em]
event void e,f;
loop do
    par/or do
        await e;
    with
        emit e;     // w/o delayed awaits, the emit awakes 1st trail
        await f;    // and restarts the loop instantaneously
    end
end
\end{lstlisting}
\caption{ Delayed awaits prevents re-execution of statements by design.
\label{lst.await}
}
\end{figure}

In atypical scenarios requiring immediate awake, delayed awaits can be 
circumvented by manually copying or transforming the code to execute on awake.
%
For instance, sometimes we need to execute a block of code immediately, and 
then periodically, from internal event requests, as illustrated in 
Figure~\ref{lst.delay}.
In this case, the \code{await} moved to the end of the loop (ln. 10) makes the 
periodic code to also execute immediately (ln. 9), and then in reactions to 
each \code{emit} request (ln. 5).
If the periodic \code{emit} depends on a condition, then the code 
transformation becomes more intricate, requiring an extra condition test around 
the periodic code to prevent its immediate execution.
%
On the one hand, we transfer the burden of dealing with these corner cases to 
the programmer.
On the other hand, we simplify the semantics of the language and eliminate the 
need for analysis to deal with schizophrenic statements.

\begin{figure}[t]
%\begin{minipage}[t]{0.50\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3em]
event void e;
par do
    loop do
        <...>     // code that awaits some period
        emit e;   // periodic request
    end
with
    loop do
        <...>     // code to execute immediately and then periodically
        await e;  // await after
    end
end
\end{lstlisting}
%\end{minipage}
%
%\begin{minipage}[t]{0.50\linewidth}
%\begin{lstlisting}[numbers=left,xleftmargin=3em]
%event void e;
%var bool should_execute = false;
%par do
%    loop do
%        <...>         // code that awaits
%        if <...> then // some conditon
%            should_execute = true;
%            emit e;
%        end
%    end
%with
%    loop do
%        if should_execute then
%            <...>     // code to execute
%        end
%        await e;
%    end
%end
%\end{lstlisting}
%\end{minipage}
%
\caption{ An example that circumvents the \emph{delayed await} restriction by 
post-fixing the \code{await} inside the \code{loop}.
% (in the left), and by copying the condition test (in the right).
\label{lst.delay}
}
\end{figure}

@@

Embedded applications make extensive use of global memory and shared resources, 
such as through memory-mapped registers and system calls to drivers.
Hence, an important goal of \CEU is to ensure a reliable behavior for programs 
with concurrent lines of execution sharing memory and interacting with the 
environment through system calls.

Esterel is only deterministic with respect to reactive control: ``the same 
sequence of inputs always produces the same sequence of 
outputs''~\cite{esterel.primer}.
%
However, the execution order for operations within a reaction is 
non-deterministic: ``if there is no control dependency, as in \code{<<call f1() 
|| call f2()>>}, the order is unspecified and it would be an error to rely on 
it''~\cite{esterel.primer}.
%
For this reason, Esterel assumes that external calls to the host language do 
not perform side effects.
%
Therefore, Esterel syntactically forbids sharing memory between lines of 
execution to preserve determinism: \emph{``if a variable is written by some 
thread, then it can neither be read nor be written by concurrent 
threads''}~\cite{esterel.primer}.

Concurrency in \CEU is characterized when two or more trail segments in 
parallel execute during the same reaction.
A trail segment is a sequence of statements followed by an \code{await} (or 
termination).
%
In the program in the left of Figure~\ref{lst.shared}, the two assignments to 
\code{x} (ln. 5,8) can never run concurrently, because each trail segment 
reacts to different input events (ln. 4,7), which cannot occur simultaneously 
(according to Section~\ref{sec.ceu.ints}).
%
However, the program in the right is non-deterministic, because the two 
assignments to \code{y} (ln. 5,8) occur in the same reaction to input \code{A} 
(ln. 4,7).

\begin{figure}
\begin{minipage}[t]{0.50\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3em]
input void A, B;
var int x = 1;
par/and do
    await A;
    x = x + 1;
with
    await B;
    x = x * 2;
end
\end{lstlisting}
\end{minipage}
%
\begin{minipage}[t]{0.50\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3em]
input void A;
var int y = 1;
par/and do
    await A;
    y = y + 1;
with
    await A;
    y = y * 2;
end
\end{lstlisting}
\end{minipage}
%\rule{8.4cm}{0.37pt}
\caption{ Shared-memory concurrency in \CEU:
the code in the left is safe because the trails access \code{x} atomically in 
different reactions;
the code in the right is unsafe because both trails access \code{y} in the same 
reaction.
\label{lst.shared}
}
\end{figure}

\CEU performs a temporal analysis at compile time and detects concurrent 
accesses to shared variables, as follows:
if a variable is written in a trail segment, then a concurrent trail segment 
cannot read or write to that variable, nor dereference a pointer of that 
variable type.
An analogous policy is applied for pointers \emph{vs} variables and pointers 
\emph{vs} pointers, as well as for system calls with side effects (e.g., 
\code{printf}).%
\footnote{\CEU assumes all system calls perform side effects, unless they are 
annotated as ``\texttt{@pure}''.}
%The algorithm for the analysis holds the set of all events in preceding 
%\code{await} statements for each variable access.
%Then, the sets for all accesses in parallel trails are compared to assert that 
%no events are shared among them.
%Otherwise the compiler warns about the suspicious accesses.

@@

\begin{figure}[b]
\begin{minipage}[t]{0.48\linewidth}
\begin{lstlisting}[mathescape=true]
// ESTEREL
input A;
[
    await A;
    call f1();
$\|$
    await A;
    call f2();
];
\end{lstlisting}
\end{minipage}
%
\begin{minipage}[t]{0.48\linewidth}
\begin{lstlisting}
// CEU
input void A;
par/and do
    await A;
    _f1();
with
    await A;
    _f2();
end
\end{lstlisting}
\end{minipage}
%\rule{8.4cm}{0.37pt}
\caption{ In Esterel, the execution order between \code{f1} and \code{f2} is 
unspecified, whereas in \CEU, \code{\_f1} executes before \code{\_f2} due to 
deterministic scheduling based on lexical order.
\label{lst.det}
}
\end{figure}

@@

To illustrate the risks related to abortion, consider the unsafe example in the 
left of Figure~\ref{lst.fin}, which does not compile in \CEU.
It describes the state machine of the data collection protocol \emph{CTP} for 
sensor networks~\cite{wsn.ctp}.
%
The input and output events represent the external interface of the protocol 
(ln. 1--2).
%
The protocol has to transmit a packet every minute (with \code{SEND\_ENQUEUE}), 
unless it receives a \code{RETRANSMIT} request to immediately re-transmit it, 
or a \code{STOP} request to terminate.
%
The protocol is implemented with two main trails:
one simply monitors the stopping event (ln. 4);
the other periodically transmits the packet (ln. 6--19).
%
The periodic transmission is a loop that starts two other trails (ln. 7--18):
one handles the immediate retransmission request (ln. 8);
the other transmits the packet and waits for a confirmation (ln. 10--17).
%
%
The actual transmission (ln. 13--16) is enclosed with a \code{par/and} that 
takes at least one minute before looping (ln. 11), in accordance with the 
specification.
Note that the \code{emit SEND\_ENQUEUE} (ln. 15) is \emph{asynchronous}, 
handing to the radio driver a pointer to the lexically-scoped packet (ln. 13).
The driver makes the transmission in the background, holding the packet until 
it signals the application with a \code{SEND\_ACK} event to acknowledge 
completion (ln. 16).
%
At any time, the client may request a retransmission (ln. 8), which terminates 
the \code{par/or} (ln. 7) and restarts the loop (ln. 6), but does not abort the 
ongoing transmission initiated with the \code{emit SEND\_ENQUEUE} (ln. 15).
%
The client may also request to stop the whole protocol at any time (ln. 4).
%
Therefore, if the sending trail is aborted by the \code{STOP} or 
\code{RETRANSMIT} requests, the packet buffer goes out of scope (ln. 13), 
leaving behind a \emph{dangling pointer} in the radio driver, which will 
possibly transmit corrupted data.
%
%\subsubsection{Finalization}
%\label{sec.ceu.fin}

The unsafe example in the left of Figure~\ref{lst.fin} does not compile because 
\CEU tracks the interaction of \code{par/or} compositions with local variables 
and stateful output events calls in order to preserve safe abortion of trails.
%
%Finalization clauses are fundamental to preserve the orthogonality of 
%\code{par/or} compositions in SSRP.
%
\CEU enforces a \emph{finalization} clause to accompany the output request.
The code in the right of Figure~\ref{lst.fin} properly cancels the packet 
transmission when the block of \code{buffer} goes out of scope, i.e., the 
finalization clause (after the \code{with}, ln. 18) executes automatically on 
external abortion.%
\footnote{
Note that the compiler only enforces the programmer to write the finalization 
clause, but cannot check if it actually handles the resource properly.
}
%
%Note also that \CEU environments rely on \C libraries that only provide 
%asynchronous I/O and non-blocking functions~\cite{ceu.sensys13}.

@@

\begin{itemize}
\item The program is awaiting in all trails, i.e., function $pop$ returns 
$(0,\{\})$.
\item The program terminates, i.e., the small-step rules transform the whole 
program into a $nop$.
\end{itemize}
%
In Section~\ref{XXX} we show that by imposing syntactic restrictions to 
programs, reaction chains always reach one of these conditions in a finite 
number of steps, meaning that reactions to the environment always execute in 
bounded time.

@@

\subsection{Interaction with the Environment}
